---
title: "RUHU movements - Data prep"
author: "Gannon et al."
date: "May 2021"
output:
  github_document:
    pandoc_args: --webtex
bibliography: /Users/dusty/Documents/zotero_library.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)

  require(tidyverse)
  require(sf)
  require(raster)
  require(lubridate)
  require(here)

# User-defined functions

## Find which elements of a vector are not in another
  `%notin%` <- Negate(`%in%`)
  
## length with null values
  length2 <- function(x){
    if(is.null(x)){
      return(0)
    } else{
      return(length(x))
    }
  }

```

## Overview
 The final dataset will include the frequency of hummingbird movements among different RFID readers across the HJA meadows in each of four years. We combine these movements with information on distances between readers, cover-type in which each reader was placed, and the proportion of intervening forest, as well as the number of birds using each feeder.
 
 
#### Visit data
  
  The raw data have already been filtered down to records that constitute "separate" visits (reads separated by at least 30s or reads of different birds). To filter the raw read data down to the visit data, see the [reads_to_visits markdown file](). The visit data include six columns of data. The fields include:
  
* PIT tag ID (character string)

* Date (yyyy-mm-dd) of visit

* Time (hh:mm:ss) of visit

* Seconds past midnight (integer)

* Year $\in$ {2014,2015,2016,2017}

* Reader-feeder site (Complex_reader)

$~$


```{r data}
#read in data
   vis <- read.table(here("Data","all_visits.txt"), 
                   header = T, sep = "\t", as.is = T)
   
# add complex id
 split <- as.data.frame(str_split(vis$site, "_", simplify = T))
 vis <- cbind(vis[,c(1:4)], split)
   
 names(vis) <- c("bird", "date", "time", "seconds", "complex", "reader")

# combine C-upper and C-lower with C
  # These readers are in the same meadow, 
  # so we can consider visits as evidence the bird is in that meadow
  vis$reader[which(vis$reader == "CU")] <- "C"
  vis$reader[which(vis$reader == "CL")] <- "C"
  vis$reader[which(vis$reader == "CW")] <- "C"
  vis$reader[which(vis$reader == "CE")] <- "C"
   
# combine complex and reader to make unique identifier for each reader
  vis$reader <- paste(vis$complex, vis$reader, sep = "_")

# load data on reader locations 
  rlocs <- read.table(here("Data", "hja_reader_locations_centercomb.txt"), 
                      header = T, sep = "\t", as.is = T)
 
```

$~$


**Convert records to movements**
 
 To, create a movement dataset, sort by `bird` then `date` then `time`. Loop through to ask whether a bird was recorded at reader $i$ at time $t$, then again at reader $j$ at time $t'>t$ within the same day.
 
```{r movements}
# order the dataframe
  vis_sort <- vis[order(vis$bird, vis$date, vis$seconds), ]

# add a column for the year
  vis_sort$year <- str_extract(vis_sort$date, "[[:digit:]]{4}")
  
# turn date into day of the year
  vis_sort$date2 <- ymd(vis_sort$date)
  vis_sort$day <- yday(vis_sort$date)

# create variables for use later
  rdrs <- rlocs$reader
  nrdrs <- length(rdrs)
  
```


**Loop to create movement data**

```{r}  

# create movement dataset
  birds_all <- unique(vis_sort$bird)
  yrs <- sort(unique(vis_sort$year))
  mv_columns <- c("reader1", "reader2", "year", "count", "birds1", "birds2")
  mv <- matrix(nrow = 1, ncol = length(mv_columns))
  colnames(mv) <- mv_columns
  mv <- as.data.frame(mv)
  
# create running list of birds using each feeder in each year
  birds_at_feeders_cnt <- matrix(
    data = 0,
    nrow = nrdrs,
    ncol = length(yrs)
  )
  rownames(birds_at_feeders) <- rdrs
  colnames(birds_at_feeders) <- yrs
  
  birds_at_feeders <- tibble(
    rdr = rdrs,
    birds_2014 = vector(mode = "list", length = nrdrs),
    birds_2015 = vector(mode= "list", length = nrdrs),
    birds_2016 = vector(mode = "list", length = nrdrs),
    birds_2017 = vector(mode = "list", length = nrdrs)
  )

# Fill dataframe with movements
  for(y in yrs){
    df_temp <- subset(vis_sort, year==y)
    df_temp <- df_temp[order(df_temp$bird, df_temp$day, df_temp$seconds), ]
    ds <- 0
  # create a count for each possible movement
  # for each year
    cnts <- matrix(data=0, nrow = nrdrs, ncol = nrdrs)
     if(nrow(df_temp) > 1){
      for(i in 2:nrow(df_temp)){
        if((df_temp$date[i] == df_temp$date[i-1]) &
           (df_temp$bird[i] == df_temp$bird[i-1]) &
            (df_temp$reader[i] != df_temp$reader[i-1])){
          # if all conditions hold, find reader numbers
            r1 <- which(rdrs == df_temp$reader[i-1])
            r2 <- which(rdrs == df_temp$reader[i])
          #add a count to that cell  
            cnts[r1,r2] <- cnts[r1,r2] + 1
          # add bird counts
            yr <- which(yrs == y)
            if(
              df_temp$bird[i] %notin%
              birds_at_feeders[[yr+1]][[r1]]
            ){
            birds_at_feeders[[yr+1]][[r1]] <- 
              c(
                birds_at_feeders[[yr+1]][[r1]],
                df_temp$bird[i]
              )
            }
            if(
              df_temp$bird[i] %notin%
              birds_at_feeders[[yr+1]][[r2]]
            ){
              birds_at_feeders[[yr+1]][[r2]] <- 
              c(
                birds_at_feeders[[yr+1]][[r2]],
                df_temp$bird[i]
              )
            }
        }
      }
     }
      # get counts of birds at feeders
      bird_cnts <- map_dbl(as.list(birds_at_feeders[[yr+1]]),
                           ~length(.x))
      # combine with movement dataframe
      if(y=="2014"){
        rems <- which(rlocs$meadow_complex == "CM" | rlocs$meadow_complex == "LOM")
        cnts_y <- cnts[-rems,-rems]
        rdrs_y <- rdrs[-rems]
        brdcnts_y <- bird_cnts[-rems]
      } else if(y=="2015"){
        rems <- which(rlocs$meadow_complex == "CM")
        rdrs_y <- rdrs[-rems]
        cnts_y <- cnts[-rems,-rems]
        brdcnts_y <- bird_cnts[-rems]
      } else{
        rdrs_y <- rdrs
        cnts_y <- cnts
        brdcnts_y <- bird_cnts
      }
        df_temp2 <- data.frame(reader1=rep(rdrs_y, length(rdrs_y)),
                               reader2=rep(rdrs_y, each=length(rdrs_y)),
                               year=rep(y, length(rdrs_y)^2),
                               count=as.vector(cnts_y),
                               birds1=rep(brdcnts_y, length(rdrs_y)),
                               birds2=rep(brdcnts_y, each=length(rdrs_y)))
        mv <- rbind(mv, df_temp2)
    
  }
  
  mv <- mv[-1,]
      
```


**Remove rows for $i \to i$ movements.**

```{r}

  rows_ii <- which(mv$reader1==mv$reader2)

  mv <- mv[-rows_ii, ]

```


## Adding in landscape variables

 Now add in spatial data pertaining to each "edge": distance between readers, proportion of intervening forest, and cover types connected by the edge (i.e. meadow-meadow, meadow-mix, meadow-forest, meadow-scrub).
 
**Edge length**
```{r edge distances}
 # coordinate matrix for reader locations
  coords_r <- as.matrix(rlocs[,c("x","y")])
  rownames(coords_r) <- rlocs$reader
  ones <- matrix(1, nrow = nrow(coords_r), ncol = 1)

# compute squared euclidean distance matrix
  D_rsq <- diag(coords_r%*%t(coords_r))%*%t(ones) + 
    ones%*%t(diag(coords_r%*%t(coords_r))) - 2*(coords_r%*%t(coords_r))

# square root of values and label the rows and cols
  D_r <- sqrt(D_rsq)
  rownames(D_r) <- rlocs$reader
  colnames(D_r) <- rownames(D_r)
 
# vectorize the distance matrix 
  dist_list <- data.frame(reader1=rep(rdrs,nrdrs),
                          reader2=rep(rdrs, each=nrdrs),
                          dist.km=as.vector(D_r)/1000)
# add distance to movement data
  mv$edgename <- paste(mv$reader1, mv$reader2, sep = "-")
  dist_list$edgename <- paste(dist_list$reader1, dist_list$reader2, sep = "-")
  mv_wdist <- merge(mv, dist_list[,c(3,4)], sort=F)

```

$~$

**Cover types**

```{r cover types}
#habitat for reader 1
  mv_wdist <- merge(mv_wdist, rlocs[,c(1,6)], by.x="reader1", by.y="reader")
  names(mv_wdist)[which(names(mv_wdist)=="habitat")] <- "hab1"
  
#habitat for reader 2
  mv_wdist <- merge(mv_wdist, rlocs[,c(1,6)], by.x="reader2", by.y="reader")
  names(mv_wdist)[which(names(mv_wdist)=="habitat")] <- "hab2"
  
# connectivity reader 1
  mv_wdist <- merge(mv_wdist, rlocs[,c(1,10)], by.x="reader1", by.y="reader")
  names(mv_wdist)[which(names(mv_wdist) == "prop_for100")] <- "p_for1"
  
# connectivity reader 2
  mv_wdist <- merge(mv_wdist, rlocs[,c(1,10)], by.x="reader2", by.y="reader")
  names(mv_wdist)[which(names(mv_wdist) == "prop_for100")] <- "p_for2" 

```

$~$

**Proportion of forest vs. non-forest cells intersected by each edge**

```{r proportion forest}

# Create points data
 utm_proj <- CRS("+proj=utm +zone=10 +ellps=GRS80 +datum=NAD83 +units=m +no_defs")
 crs <- st_crs(utm_proj)
 
# create list to store start-end coords
  edges_coordlist <- vector(mode = "list")

  for(i in 1:nrdrs){
      for(j in 1:nrdrs){
        coords <- as.matrix(rbind(coords_r[i, ],
                                coords_r[j, ]))
        nm <- paste(rownames(coords_r)[i], 
                    rownames(coords_r)[j],
                    sep = "-")
        edges_coordlist[[nm]] <- coords
      }
    }
  
# convert to lines
  edges_aslines <- map(edges_coordlist, ~Line(.))
 
# Convert the many lines to a single lines object
  edges_ls <- map(1:length(edges_aslines), 
                          ~Lines(edges_aslines[[.]], ID=names(edges_aslines)[.]))
  
# create one spatial lines object
  edges_spl <- SpatialLines(edges_ls, proj4string = utm_proj)
  
# read in forest data
  forest <- raster(here("Data","forest.tif"))
  
# extract proportion of forest between the nodes
  beginCluster(n=2)
    prop_for <- raster::extract(forest, edges_spl, fun=mean, na.rm=T,
                      df=T, buffer=25)
  endCluster()  
  prop_for$ID <- names(edges_aslines)

  
#merge two datasets
   mv_wpropfor <- merge(mv_wdist, prop_for,
                        by.x="edgename",
                        by.y="ID")
# reorder
   cols.order <- c("reader1", "reader2", "year", "hab1", "hab2", 
                   "p_for1", "p_for2", "dist.km", "forest", "count",
                   "birds1", "birds2")
   mv_noedgename <- mv_wpropfor[,cols.order]
   
  names(mv_noedgename) <- c("reader_i", "reader_j", "year", "hab_i", "hab_j",
                  "pfor_100m_i", "pfor_100m_j", "dist_ij", "pfor_ij", "count",
                  "nbirds_i", "nbirds_j")
  
# now create a dyad identifier
  mv_noedgename$dyad <- NA
  for(i in 1:nrow(mv_noedgename)){
    dyad_i <- sort(c(mv_noedgename$reader_i[i], mv_noedgename$reader_j[i]))
    mv_noedgename$dyad[i] <- paste(dyad_i, collapse = '-')
  }
  
  mv_noedgename <- mv_noedgename[,c(
    1,2, ncol(mv_noedgename),
    3:(ncol(mv_noedgename)-1)
  )]

```


**Including an offset for the number of days the readers were maintained on the landscape for a given year**


```{r}

  active_days <- data.frame(year=c(2014:2017),
                            active_days = c(11,64,54,50))

  mv_final <- merge(mv_noedgename, active_days)

```

### Write out the cleaned movement data

```{r final undirected data}

  mv_final <- mv_final[order(mv_final$year, mv_final$reader_i, mv_final$reader_j),]
 
  write.table(mv_final, 
              file = here("Data", "total_movements_by_year.tsv"), 
              row.names = F, 
              sep = "\t",
              quote = F)

```




















