---
title: "Sender-receiver Poisson model fit"
author: "D. G. Gannon"
date: "May 2021"
output:
  github_document:
    pandoc_args: --webtex
bibliography: /Users/dusty/Documents/zotero_library.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#packages
  require(rstan)
  require(loo)
  require(igraph)
  require(raster)
  require(ggplot2)
  require(gridExtra)
  require(here)
  require(tidyverse)
  


# function to fill square matrix by rows and skip the diagonal

 fill_Amat <- function(x,dim){
  A_nodi <- matrix(x, nrow = dim, ncol = dim-1, byrow = T)
  A <- matrix(nrow = dim, ncol = dim)
  for(i in 1:nrow(A_nodi)){
    if(i>1){
     for(j in 1:(i-1)){
       A[i,j] <- A_nodi[i,j]
     }
    }
    if(i+1 <= dim){
      for(k in (i+1):dim){
        A[i,k] <- A_nodi[i,(k-1)]
      }
   }
  }
  return(A)
 }
 

```

Let $y_{i,j,k} \in \mathbb{N}^+$ be the number of movements detected between readers $i$ and $j$ by hummingbirds in year $k$, where $i \ne j = 1,2,...,R$. Following @hoff2005b, we model the number of movements in a bilinear model:

$$
 \log (\lambda_{i,j,k}) =  {\bf x}_{ij}\boldsymbol \beta + u_i + w_j + \gamma_{ij} + \eta_k +\log(b_k) + \log(d_k)
$$
where ${\boldsymbol \beta}$ is a vector of dyad specific effects with ${\bf x}_{i,j}$ the vector of dyad-specific regressors for dyad $\{i,j\}$. The effects $u_i$ and $w_i$ are the average effects of reader $i$ as a sender and receiver (respectively), and the term $\gamma_{i,j}$ is an average effect on movement for the pair of readers $\{i,j\}$. Finally, $b_k$ and $d_k$ are offsets for the cumulative number of birds that had been implanted with RFID tags in year $k$ and the number of days the readers were maintained in year $k$ (respectively).

To induce dependence among observations that involve reader $i$, we assume

$$
\begin{aligned}
\begin{bmatrix}
u_i \\
w_i
\end{bmatrix} &\sim \mathcal{N}\left({\bf 0},\ {\boldsymbol \Sigma}_{uw} \right),\\
{\boldsymbol \Sigma}_{uw} &= \begin{bmatrix}
\sigma_{u}^2 & \sigma_{uw}\\
\sigma_{uw} & \sigma_w^2
\end{bmatrix},
\end{aligned}
$$

such that observations that have a common sender or receiver may be correlated. Importantly, this allows for the potential that good senders may be poor receivers, so negative correlation is a possibility. Finally, let

$$
\begin{aligned}
\begin{bmatrix}
\gamma_{i,j}\\
\gamma_{j,i}
\end{bmatrix} &\sim \mathcal{N}({\bf 0},\ {\boldsymbol \Sigma}_\gamma),\\
{\boldsymbol \Sigma}_\gamma &= \begin{bmatrix}
\sigma_\gamma^2 & \rho\sigma_\gamma^2\\
\rho\sigma_\gamma^2 & \sigma_\gamma^2
\end{bmatrix},
\end{aligned}
$$

where $\rho$ is the correlation between the rate of movement from $i\to j$ and $j \to i$. These differ from standard random effects models because they allow for negative correlation of the observations within a dyad.

## Stan Model Code

```{stan output.var="soc_net_poisson", eval=FALSE}

data{

  int<lower=0> N;             //number of observations (edges*years)
  int<lower=0> R;             //number of nodes (readers)
  int<lower=0> D;             //number of dyads
  int<lower=0> P;             //number of fixed effects
  
  int<lower=0,upper=R> r1[N];     //reader 1 index
  int<lower=0,upper=R> r2[N];     //reader 2 index
  int<lower=0,upper=D> dyad[N];   //index for the dyad
  int<lower=0> order[N];          //order for the dyad
  
  matrix[N,P] X;                  //fixed effects model matrix
  real<lower=0> os_d[N];          //number of days connections were active
  real<lower=0> os_b[N];          //cumulative number of birds with RFID tags
  int<lower=0> y[N];              //number of movements

}

parameters{

  vector[P] beta;                //regression parameters
  vector[2] u_raw[R];            //random effects for sender/receiver
  vector[2] gamma_raw[D];        //random effects for the dyad

  vector<lower=0>[2] sigma_u;            //scale for sender/receiver effects
  cholesky_factor_corr[2] L_omega_u;
  
  real<lower=0> sigma_gamma;        //scale for dyad effects
  cholesky_factor_corr[2] L_omega_gamma;
  

}

transformed parameters{

  vector[2] u[R];                
  vector[2] gamma[D];
  real<lower=0> lambda[N];
  
  for(r in 1:R){
    u[r] = diag_pre_multiply(sigma_u, L_omega_u)*u_raw[r];
  }
  
  for(d in 1:D){
    gamma[d] = diag_pre_multiply(rep_vector(sigma_gamma,2),
                  L_omega_gamma)*gamma_raw[d];
  }
  
  //create linear predictor
  for(i in 1:N){
    lambda[i] = exp(X[i,]*beta + u[r1[i],1] + u[r2[i],2] +
                    gamma[dyad[i],order[i]] 
                    + log(os_d[i]) + log(os_b[i]));
  }
  
}

model{

  //priors
  beta ~ normal(0,1);
  sigma_u ~ normal(0,2);
  sigma_gamma ~ normal(0,2);
  L_omega_u ~ lkj_corr_cholesky(5);
  L_omega_gamma ~ lkj_corr_cholesky(5);
  
  
  //model
  for(r in 1:R){
    u_raw[r] ~ std_normal();
  }

  for(d in 1:D){
    gamma_raw[d] ~ std_normal(); 
  }
  
  
  for(i in 1:N){
    y[i] ~ poisson(lambda[i]);
  }


}

generated quantities{

  vector[N] log_lik;
  vector[N] y_hat;
  matrix[2,2] omega_u;
  matrix[2,2] omega_gamma;
  
  omega_u = L_omega_u*L_omega_u';
  omega_gamma = L_omega_gamma*L_omega_gamma';
  
  for(i in 1:N){
    log_lik[i] = poisson_lpmf(y[i] | lambda[i]);
    y_hat[i] = poisson_rng(lambda[i]);
  }

}


```

### Load data

```{r}

# number of movements for each year
  mvmnts <- read.table(file = here("Data","total_movements_by_year.tsv"), 
                       header = T, as.is = T, sep = "\t")
  
# reader locations
  rlocs <- read.table(file = here("Data", "hja_reader_locations_centercomb.txt"),
                      header = T, sep = "\t")
  rdrs <- rlocs$reader
  
# bird capture data
  caps <- read_csv(file = here("Data", "hja_hummingbird_captures_rfid.csv"))
  birds_yr <- group_by(caps, yr) %>%
    summarise(n=n())
  
  birds_yr$cum_birds <- cumsum(birds_yr$n)
  
# merge back with movement data
  mvmnts <- merge(mvmnts, birds_yr, by.x="year", by.y="yr")

# create a column for the order
  #if the position for reader 1 > reader 2, order=1
 mvmnts$order <- NA
 for(i in 1:nrow(mvmnts)){
   r1 <- which(rdrs == mvmnts$reader_i[i])
   r2 <- which(rdrs == mvmnts$reader_j[i])
   if(r1 < r2){
     mvmnts$order[i] <- 1
   } else{
     mvmnts$order[i] <- 2
   }
 }
 
# indicator variables for whether a reader is inside the forest
 mvmnts$forest_i <- 0
 mvmnts$forest_i[which(mvmnts$hab_i == "forest")] <- 1
 
 mvmnts$forest_j <- 0
 mvmnts$forest_j[which(mvmnts$hab_j == "forest")] <- 1
 
 mvmnts$forest_ij <- with(mvmnts,
                          (forest_i + forest_j)/2)
 
```

# Compile data
```{r compile data}

  N <- nrow(mvmnts)
  y <- mvmnts$count
  X <- model.matrix(~ dist_ij + pfor_ij + forest_ij, data = mvmnts)
  
  P <- dim(X)[2]
  ord <- mvmnts$order
  dyad <- as.integer(as.factor(mvmnts$dyad))
  D <- length(unique(dyad))
  
  r1 <- as.integer(factor(mvmnts$reader_i, levels = rdrs))
  r2 <- as.integer(factor(mvmnts$reader_j, levels = rdrs))
  R <- length(unique(r1))
  os_d <- (mvmnts$active_days)/7
  os_b <- mvmnts$cum_birds
  
# compile data into list
  mod.data <- list(N=N, P=P, R=R, D=D, r1=r1,
                   r2=r2, dyad=dyad, order=ord,
                   X=X, os_d=os_d, os_b=os_b,
                   y=y)
```

# Fitting the model
```{r eval=FALSE}
# fit the model
  fit <- sampling(soc_net_poisson, data=mod.data, chains=2, 
                  control=list(adapt_delta=0.9), init=0)

  
  

```

```{r done, echo=FALSE}
knitr::knit_exit()
```


```{r load fit, echo=FALSE}
load( here("Data", "fit_sender_receiver.RData"))
```


```{r}

 loglik1 <- extract_log_lik(fit, merge_chains = F)
 reff1 <- relative_eff(exp(loglik1))
 fit_loo1 <- loo(loglik1, r_eff = reff1)
 
```

```{r save fit, echo=FALSE, eval=FALSE}

 save(fit, fit_loo1, mod.data,
      file = here("Data", "fit_sender_receiver.RData"))

```









