---
title: "Sender-receiver Poisson model fit"
author: "D. G. Gannon"
date: "October 2020"
output:
  github_document:
    pandoc_args: --webtex
bibliography: /Users/dusty/Documents/zotero_library.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#packages
  require(rstan)
  require(loo)
  require(igraph)
  require(raster)
  require(ggplot2)
  require(gridExtra)
  require(here)
  require(tidyverse)
  


# function to fill square matrix by rows and skip the diagonal

 fill_Amat <- function(x,dim){
  A_nodi <- matrix(x, nrow = dim, ncol = dim-1, byrow = T)
  A <- matrix(nrow = dim, ncol = dim)
  for(i in 1:nrow(A_nodi)){
    if(i>1){
     for(j in 1:(i-1)){
       A[i,j] <- A_nodi[i,j]
     }
    }
    if(i+1 <= dim){
      for(k in (i+1):dim){
        A[i,k] <- A_nodi[i,(k-1)]
      }
   }
  }
  return(A)
 }
 
 check_energy <- function(fit, quiet=FALSE) {
  sampler_params <- get_sampler_params(fit, inc_warmup=FALSE)
  no_warning <- TRUE
  for (n in 1:length(sampler_params)) {
    energies = sampler_params[n][[1]][,'energy__']
    numer = sum(diff(energies)**2) / length(energies)
    denom = var(energies)
    if (numer / denom < 0.2) {
      if (!quiet) print(sprintf('Chain %s: E-FMI = %s', n, numer / denom))
      no_warning <- FALSE
    }
  }
  if (no_warning) {
    if (!quiet) print('E-FMI indicated no pathological behavior')
    if (quiet) return(TRUE)
  } else {
    if (!quiet) print('  E-FMI below 0.2 indicates you may need to reparameterize your model')
    if (quiet) return(FALSE)
  }
}

```

Let $y_{i,j,k} \in \mathbb{N}^+$ be the number of movements detected between readers $i$ and $j$ by hummingbirds in year $k$, where $i \ne j = 1,2,...,R$. Following @hoff2005b, we model the number of movements in a bilinear model:

$$
 \log (\lambda_{i,j,k}) =  {\bf x}_{ij}\boldsymbol \beta + u_i + w_j + \gamma_{ij} + \eta_k +\log(b_k) + \log(d_k)
$$
where ${\boldsymbol \beta}$ is a vector of dyad specific effects with ${\bf x}_{i,j}$ the vector of dyad-specific regressors for dyad $\{i,j\}$. The effects $u_i$ and $w_i$ are the average effects of reader $i$ as a sender and receiver (respectively), and the term $\gamma_{i,j}$ is an average effect on movement for the pair of readers $\{i,j\}$. Finally, $b_k$ and $d_k$ are offsets for the cumulative number of birds that had been implanted with RFID tags in year $k$ and the number of days the readers were maintained in year $k$ (respectively).

To induce dependence among observations that involve reader $i$, we assume

$$
\begin{split}
\begin{bmatrix}
u_i \\
w_i
\end{bmatrix} \sim \mathcal{N}\left({\bf 0},\ {\boldsymbol \Sigma}_{uw} \right),\\
{\boldsymbol \Sigma}_{uw} = \begin{bmatrix}
\sigma_{u}^2 & \sigma_{uw}\\
\sigma_{uw} & \sigma_w^2
\end{bmatrix},
\end{split}
$$

such that observations that have a common sender or receiver may be correlated. Importantly, this allows for the potential that good senders may be poor receivers, so negative correlation is a possibility. Finally, let

$$
\begin{split}
\begin{bmatrix}
\gamma_{i,j}\\
\gamma_{j,i}
\end{bmatrix} \sim \mathcal{N}({\bf 0},\ {\boldsymbol \Sigma}_\gamma),\\
{\boldsymbol \Sigma}_\gamma = \begin{bmatrix}
\sigma_\gamma^2 & \rho\sigma_\gamma^2\\
\rho\sigma_\gamma^2 & \sigma_\gamma^2
\end{bmatrix},
\end{split}
$$

where $\rho$ is the correlation between the rate of movement from $i\to j$ and $j \to i$. These differ from standard random effects models because they allow for negative correlation of the observations within a dyad.



```{stan output.var="soc_net_poisson", eval=FALSE}

data{

  int<lower=0> N;             //number of observations (edges*years)
  int<lower=0> R;             //number of nodes (readers)
  int<lower=0> D;             //number of dyads
  int<lower=0> P;             //number of fixed effects
  
  int<lower=0,upper=R> r1[N];     //reader 1 index
  int<lower=0,upper=R> r2[N];     //reader 2 index
  int<lower=0,upper=D> dyad[N];   //index for the dyad
  int<lower=0> order[N];          //order for the dyad
  
  matrix[N,P] X;                  //fixed effects model matrix
  real<lower=0> os_d[N];          //number of days connections were active
  real<lower=0> os_b[N];          //cumulative number of birds with RFID tags
  int<lower=0> y[N];              //number of movements

}

parameters{

  vector[P] beta;                //regression parameters
  vector[2] u_raw[R];            //random effects for sender/receiver
  vector[2] gamma_raw[D];        //random effects for the dyad

  vector<lower=0>[2] sigma_u;            //scale for sender/receiver effects
  cholesky_factor_corr[2] L_omega_u;
  
  real<lower=0> sigma_gamma;        //scale for dyad effects
  cholesky_factor_corr[2] L_omega_gamma;
  

}

transformed parameters{

  vector[2] u[R];                
  vector[2] gamma[D];
  real<lower=0> lambda[N];
  
  for(r in 1:R){
    u[r] = diag_pre_multiply(sigma_u, L_omega_u)*u_raw[r];
  }
  
  for(d in 1:D){
    gamma[d] = diag_pre_multiply(rep_vector(sigma_gamma,2),
                  L_omega_gamma)*gamma_raw[d];
  }
  
  //create linear predictor
  for(i in 1:N){
    lambda[i] = exp(X[i,]*beta + u[r1[i],1] + u[r2[i],2] +
                    gamma[dyad[i],order[i]] 
                    + log(os_d[i]) + log(os_b[i]));
  }
  
}

model{

  //priors
  beta ~ normal(0,1);
  sigma_u ~ normal(0,2);
  sigma_gamma ~ normal(0,2);
  L_omega_u ~ lkj_corr_cholesky(5);
  L_omega_gamma ~ lkj_corr_cholesky(5);
  
  
  //model
  for(r in 1:R){
    u_raw[r] ~ std_normal();
  }

  for(d in 1:D){
    gamma_raw[d] ~ std_normal(); 
  }
  
  
  for(i in 1:N){
    y[i] ~ poisson(lambda[i]);
  }


}

generated quantities{

  vector[N] log_lik;
  vector[N] y_hat;
  matrix[2,2] omega_u;
  matrix[2,2] omega_gamma;
  
  omega_u = L_omega_u*L_omega_u';
  omega_gamma = L_omega_gamma*L_omega_gamma';
  
  for(i in 1:N){
    log_lik[i] = poisson_lpmf(y[i] | lambda[i]);
    y_hat[i] = poisson_rng(lambda[i]);
  }

}


```


```{r}

# number of movements for each year
  mvmnts <- read.table(file = here("Data","total_movements_by_year.tsv"), 
                       header = T, as.is = T, sep = "\t")
  
# reader locations
  rlocs <- read.table(file = here("Data", "hja_reader_locations_centercomb.txt"),
                      header = T, sep = "\t")
  rdrs <- rlocs$reader
  
# bird capture data
  caps <- read_csv(file = here("Data", "hja_hummingbird_captures_rfid.csv"))
  birds_yr <- group_by(caps, yr) %>%
    summarise(n=n())
  
  birds_yr$cum_birds <- cumsum(birds_yr$n)
  
# merge back with movement data
  mvmnts <- merge(mvmnts, birds_yr, by.x="year", by.y="yr")

# create a column for the order
  #if the position for reader 1 > reader 2, order=1
 mvmnts$order <- NA
 for(i in 1:nrow(mvmnts)){
   r1 <- which(rdrs == mvmnts$reader_i[i])
   r2 <- which(rdrs == mvmnts$reader_j[i])
   if(r1 < r2){
     mvmnts$order[i] <- 1
   } else{
     mvmnts$order[i] <- 2
   }
 }
 
# indicator variables for whether a reader is inside the forest
 mvmnts$forest_i <- 0
 mvmnts$forest_i[which(mvmnts$hab_i == "forest")] <- 1
 
 mvmnts$forest_j <- 0
 mvmnts$forest_j[which(mvmnts$hab_j == "forest")] <- 1
 
 mvmnts$forest_ij <- with(mvmnts,
                          (forest_i + forest_j)/2)
 
```


```{r model fit}

  N <- nrow(mvmnts)
  y <- mvmnts$count
  X <- model.matrix(~ dist_ij + pfor_ij + forest_ij, data = mvmnts)
  
  P <- dim(X)[2]
  ord <- mvmnts$order
  dyad <- as.integer(as.factor(mvmnts$dyad))
  D <- length(unique(dyad))
  
  r1 <- as.integer(factor(mvmnts$reader_i, levels = rdrs))
  r2 <- as.integer(factor(mvmnts$reader_j, levels = rdrs))
  R <- length(unique(r1))
  os_d <- (mvmnts$active_days)/7
  os_b <- mvmnts$cum_birds
  
# compile data into list
  mod.data <- list(N=N, P=P, R=R, D=D, r1=r1,
                   r2=r2, dyad=dyad, order=ord,
                   X=X, os_d=os_d, os_b=os_b,
                   y=y)
  
# fit the model
  fit <- sampling(soc_net_poisson, data=mod.data, chains=2, 
                  control=list(adapt_delta=0.9), init=0)

  
  

```

```{r}

 loglik1 <- extract_log_lik(fit, merge_chains = F)
 reff1 <- relative_eff(exp(loglik1))
 fit_loo1 <- loo(loglik1, r_eff = reff1)
 
```

```{r save fit, echo=FALSE, eval=FALSE}

 save(fit, fit_loo1, mod.data,
      file = here("Data", "fit_sender_receiver.RData"))

```

```{r load fit, echo=FALSE}
load( here("Data", "fit_sender_receiver.RData"))
```


```{r}

# colors for plot
 forest.col <- rgb(229,229,220, maxColorValue = 255)
 nonforest.col <- rgb(255,255,255, maxColorValue = 255)
 rdr.col <- rgb(252,163,17, maxColorValue = 255)
 edge.col <- rgb(20,33,61, maxColorValue = 255)
 # cuts <- c(-1,0.5,1.5)
 # pal <- colorRampPalette(c(nonforest.col, forest.col))
 
# load forest raster
 #forest <- raster("~/Documents/Columbine/Data/GIS_Data/forest.tif")

 y.hat <- as.data.frame(rstan::extract(fit, pars="y_hat"))

 E_y.hat <- apply(y.hat, 2, mean)
 
# create aggregated counts
 mv_counts <- data.frame(edge=paste(mvmnts$reader_i, mvmnts$reader_j, sep = "-"),
                         obs_count=mvmnts$count,
                         pred_count=E_y.hat)
 
 mvs_all <- group_by(mv_counts, edge) %>% summarise(., obs=sum(obs_count), pred=sum(pred_count))
 
# Create observed and predicted adjacency matrices
 Amat_pred <- fill_Amat(mvs_all$pred, dim=R)
 Amat_obs <- fill_Amat(mvs_all$obs, dim=R)
 
 rownames(Amat_pred) <- rlocs$reader
 colnames(Amat_pred) <- rlocs$reader
 rownames(Amat_obs) <- rlocs$reader
 colnames(Amat_obs) <- rlocs$reader
 
 g_pred <- graph_from_adjacency_matrix(Amat_pred, mode = "directed", weighted = T, diag = F)
 g_pred <- set_vertex_attr(g_pred, "x", value = rlocs$x)
 g_pred <- set_vertex_attr(g_pred, "y", value=rlocs$y)
 E(g_pred)$width <- log(E(g_pred)$weight + 1)
 E(g_pred)$arrow.size <- 0.1
 E(g_pred)$color <- edge.col
 V(g_pred)$size <- 5
 V(g_pred)$label <- NA
 V(g_pred)$color <- rdr.col
 
 
 g_obs <- graph_from_adjacency_matrix(Amat_obs, mode = "directed", weighted = T, diag = F)
 V(g_obs)$x <- rlocs$x
 V(g_obs)$y <- rlocs$y
 E(g_obs)$width <- log(E(g_obs)$weight + 1)
 E(g_obs)$arrow.size <- 0.2
 E(g_obs)$color <- edge.col
 V(g_obs)$label <- NA
 V(g_obs)$size <- 5
 V(g_obs)$color <- rdr.col
 
png(filename = here("Figures", "post_pred_fig.png"),
    height = 1800, width = 3600, res = 300)
 par(mfrow=c(1,2), mar=c(1,2,2,1))
 plot(g_obs, frame=T)
 title(main = "Observed movements", cex.main=1, adj=0)
 plot(g_pred, frame=T)
 title(main = "Posterior predictive movement rates",
       cex.main=1, adj=0)
 dev.off()
  
```

\subsection{Computing observed and predicted network metrics}

```{r node strength}

 pred.col <- rgb(229,229,220, maxColorValue = 255)
 obs.col <- rgb(252,163,17, maxColorValue = 255)
 interval.col <- rgb(20,33,61, maxColorValue = 255)

# incoming strength

  obs_instr <- apply(Amat_obs, 2, sum, na.rm=T)

  pp_instr <- matrix(nrow = nrow(y.hat), ncol = R)
  colnames(pp_instr) <- colnames(Amat_obs)
  for(i in 1:nrow(pp_instr)){
    yhat_i <- as.double(y.hat[i, ])
    Amat_i <- fill_Amat(yhat_i, dim = R)
    pp_instr[i,] <- apply(Amat_i, 2, sum, na.rm=T)
  }

  instr_df <- data.frame(reader=colnames(Amat_obs),
                         obs_instr = obs_instr/(rlocs$days_active/7),
                         mean_pred = apply(pp_instr,2,mean)/(rlocs$days_active/7),
                         low_pred=apply(pp_instr,2,quantile, probs=0.025)/(rlocs$days_active/7),
                         high_pred=apply(pp_instr,2,quantile, probs=0.975)/(rlocs$days_active/7))
  instr_df$reader <- factor(instr_df$reader, 
                            levels = colnames(Amat_obs)[order(instr_df$mean_pred)])
  
  graph_mets_theme <- theme(panel.background = element_blank(),
                            axis.line = element_line(colour = "black"),
                            axis.text.x = element_text(colour = "white"))
  
  
  instr_plot <- ggplot(data = instr_df)+
    geom_errorbar(aes(x=reader, ymin=low_pred, ymax=high_pred),
                  width=0.2, size=1.5, color=alpha(interval.col, alpha = 0.95))+
    geom_point(aes(x=reader, y=mean_pred), size=4, color="white")+
    geom_point(aes(x=reader, y=mean_pred), size=3.5, color="black")+
    geom_point(aes(x=reader,y=obs_instr), size=2, color=obs.col,
               shape=18)+
    graph_mets_theme+
    xlab("")+
    ylab("Incoming movements per week")
  
  
# Outgoing strength
  obs_outstr <- apply(Amat_obs, 1, sum, na.rm=T)

  pp_outstr <- matrix(nrow = nrow(y.hat), ncol = R)
  colnames(pp_outstr) <- rownames(Amat_obs)
  for(i in 1:nrow(pp_outstr)){
    yhat_i <- as.double(y.hat[i, ])
    Amat_i <- fill_Amat(yhat_i, dim = R)
    pp_outstr[i,] <- apply(Amat_i, 1, sum, na.rm=T)
  }

  outstr_df <- data.frame(reader=colnames(Amat_obs),
                         obs_outstr = obs_outstr/(rlocs$days_active/7),
                         mean_pred = apply(pp_outstr,2,mean)/(rlocs$days_active/7),
                         low_pred=apply(pp_outstr,2,quantile, probs=0.025)/(rlocs$days_active/7),
                         high_pred=apply(pp_outstr,2,quantile, probs=0.975)/(rlocs$days_active/7))
  outstr_df$reader <- factor(outstr_df$reader, 
                            levels = colnames(Amat_obs)[order(outstr_df$mean_pred)])
  
  outstr_plot <- ggplot(data = outstr_df)+
    geom_errorbar(aes(x=reader, ymin=low_pred, ymax=high_pred),
                  width=0.2, size=1.5, color=alpha(interval.col, alpha = 0.95))+
    geom_point(aes(x=reader, y=mean_pred), size=4, color="white")+
    geom_point(aes(x=reader, y=mean_pred), size=3.5, color="black")+
    geom_point(aes(x=reader,y=obs_outstr), size=2, color=obs.col,
               shape=18)+
    graph_mets_theme+
    xlab("")+
    ylab("Outgoing movements per week")
  
  
  grid.arrange(instr_plot, outstr_plot, nrow=2)

```









